Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-09-07 bbbq_article.tex
Better references, add percentage of TMH of SARS-CoV-2

Transmembrane helices are conserved structures that span
a cell membrane with an alpha helix.
TMHs are hydrophobic, as this is required to span the 
hydrophobic cellular lipid membrane. Additionally,
they often have a length of 23 amino acids to be able to span
the membrane.
Polypeptide fragments derived from TMHs are among the most hydrophobic,
together with the internals of soluble proteins, where the
hydrophobicity parts guide the protein to achieve its 3D configuration.
TMHs are general structures: 25 percent of the human proteome is
anchored by at least one TMH. SARS-CoV-2 has 21 TMHs, 
making up for \richel{?} percent of its entire genome.
There are multiple computational tools developed to predict which
parts of membrane proteins are TMH.

@@

One might expect that the hydrophobic epitopes presented,
for a same hydrophobicity, are as likely to stem from 
membrane proteins TMHs or from soluble proteins hydrophobic (non-TMH) regions.
For MHC-I, however, it is found that the 9-mers stemming
from TMHs are presented more often than expected by 
chance \cite{bianchi2017}
\richel{but also when compensating for hydrophobicity?}.
For MHC-II, it is unknown which percentage of binders 
is derived from TMHs for all 13-mers present in a proteome
that have the same hydrophobicity.

@@

An increased detection
is obtained by expressing a wide variety of
MHCs, which is assured by the human leukocyte antigen gene complex (HLA).
However, a different HLA will result in a different variety of MHCs,
that will display different peptide fragments. Additionally, there
may have been selection the HLA to display and or detect different
peptide fragments.

@@

In general, on would hope that evolutionary selection results in
an immune system that as most attentive for loci that are
essential for a virus, as these will be most conserved.
In SARS-CoV-2, for example, there is preliminary evidence that the strongest
selection pressure is upon residues that changes its 
virulence \cite{velazquez2020positive}.
These loci, however, only account for a small part of a pathogen's proteome.
Additionally, these essential parts differ widely between pathogens.
Because of this scarcity and variance in targets, 
one can imagine that the human immune system 
is not tailored to detect these sites, 
as hinted by upon by the aforementioned influenza study.

@@

The effective number of epitopes, however,
is much lower than the actual number of possible
epitopes, due to immunodominance. 
Immunodominance is a feature of the immune system,
due to which some epitopes dominate in causing an immune response,
where other (called subdominant) epitopes end up having no 
effect \cite{akram2012immunodominance}.
The same study mentions 20 factors that influence the strength
of an immune response.
The most relevant factor for this study is the effect of
a mutation within an epitope, 
such as the in vitro example described in \cite{berkhoff2004mutation},
where one single mutation resulted in a 5-20\% weaker
response of virus-specific CD8+ T cells \richel{Fun experiment:
do this in silico}.
Due to immunodominance, we dare assume that there can be selection
for immune system avoidance, as the effective number of loci may be
low and effects of a single mutation noticeable.

@@

When using DNA sequences, one can use a skewed rate 
in non- versus synonymous mutation
to detect the signature of selection \cite{murrell2015gene}.
Using AA sequences, however, has its advantages,
as its is closer to the actual phenotype
selection acts upon: DNA may never be translated to RNA,
or its RNA may never be transcribed \cite{diz2012proteomics}
\richel{improve upon this point}.
When using a proteome in phylogenetic research,
we know that the majority of proteins are selected to just 
maintain their function most of the time, where
is the time spans there is selection, only a few AAs
can actually increase the 'fitness' of the 
protein \cite{anisimova2009investigating}.
There, when generalizing the dynamics of mostly purifying selection (to maintain
a protein's function) and a short duration of positive selection,
those genes that are selected cannot be detected \cite{yang2000statistical}.

@@

Instead of using the membrane protein topology prediction
tool \verb;TMHMM; \cite{krogh2001predicting}, PureseqTM \cite{wang2019efficient}
is used instead.

@@

To detect the evolutionary conservation of TMHs,
we first selected all unique protein sequences per
gene from all 45k SARS-CoV-2 proteomes,
resulting in around 500 unique protein sequences per gene.
From the unique protein sequences, we made an alignment per gene,
using the msa \cite{bodenhofer2015msa} R package, 
using default settings, including the use of the default alignment tool
ClustalW \cite{thompson1994clustal}.
Per alignment, we used msa to determine the consensus sequence
and the conservation scores.
Per consensus sequences, the topology was determined using pureseqtmr.

@@

There are multiple computational tools developed to predict which
parts of membrane proteins are TMH.
In 2001, multiple tools to do so have been compared \cite{moller2001evaluation},
of which TMHMM \cite{krogh2001predicting} turned out to be the best.
TMHMM is used in the previous study \cite{bianchi2017},
yet TMHMM itself has a restrictive software license and is nearly two
decades old. 
Inspired on TMHMM, many other tools followed, such as 
Phobius (\cite{kall2004combined}),
ConPred II (\cite{arai2004conpred}),
MEMSAT3 (\cite{jones2007improving})
and MetaTM (\cite{klammer2009metatm}).
The most recent FOSS tool is called PureseqTM (\cite{wang2019efficient}),
which is used in this study.

@@

For MHC-I, there are multiple computational tools developed 
to predict epitopes. 
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCcons \cite{karosiene2012netmhccons} gave the best predictions.
A tool developed later is \verb;epitope-prediction; \cite{bianchi2017},
which uses a stabilized matrix method \cite{kim2009derivation}.
The most recent FOSS tool available is 
MHCnuggets \cite{shao2020high} (which can do both MHC-I 
and MHC-II prediction) which is used in this study

@@

Also for MHC-II, there are multiple computational tools developed 
to predict epitopes,
such as using a trained neural network (\cite{nielsen2003reliable})
or a Gibbs sampling approach (\cite{nielsen2004improved}).
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCIIpan (\cite{nielsen2008quantitative,karosiene2013netmhciipan})
gave rise to the most accurate predictions.
Later tools are \cite{zhang2013predicting},
\cite{trolle2014nettepi}, \cite{zhang2015accurate}
and the very recent MHCnuggets (\cite{shao2020high}),
the latter which is used in this study.

@@

We aimed our evolutionary experiment at TMHs, because these can
be predicted well from a protein structure,
are common structures and are present in all pathogens. 
We could have done the same experiment on beta-turn,
as also these can be predicted well \cite{petersen2010netturnp},
are common structures and are present in all pathogens.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster. 
Additionally, we would like to thank Sci-Hub \cite{himmelstein2018sci}
for allowing us to read paywalled articles while working from home.

>>>>

Transmembrane helices are structurally conserved structures that span
a cell membrane with an alpha helix.
TMHs are hydrophobic, as this is required to span the 
hydrophobic cellular lipid membrane. Additionally,
they often have a length of 23 amino acids to be able to span
the membrane.
Polypeptide fragments derived from TMHs are among the most hydrophobic,
together with the internals of soluble proteins, where the
hydrophobicity parts guide the protein to achieve its 3D configuration.
TMHs are general structures: 25 percent of the human proteome is
anchored by at least one TMH. SARS-CoV-2 has 21 TMHs, 
making up for 5.5 percent of its entire genome.
There are multiple computational tools developed to predict which
parts of membrane proteins are TMH.

@@

One might expect that the hydrophobic epitopes presented,
for a same hydrophobicity, are as likely to stem from 
membrane proteins TMHs or from soluble proteins hydrophobic (non-TMH) regions.
For MHC-I, however, it is found that the 9-mers stemming
from TMHs are presented more often than expected by 
chance (\cite{bianchi2017})
\richel{but also when compensating for hydrophobicity?}.
For MHC-II, it is unknown which percentage of binders 
is derived from TMHs for all 13-mers present in a proteome
that have the same hydrophobicity.

@@

One of many strategies to increase the number of pathogenic
peptides that can be detected is to express a wide variety of
MHCs, which is assured by the human leukocyte antigen gene complex (HLA).
However, a different HLA will result in a different variety of MHCs,
that will display different peptide fragments. Additionally, there
may have been selection the HLA to display and or detect different
peptide fragments.

@@

In general, on would hope that evolutionary selection results in
an immune system that as most attentive for loci that are
essential for a virus, as these will be most conserved.
In SARS-CoV-2, for example, there is preliminary evidence that the strongest
selection pressure is upon residues that changes its 
virulence (\cite{velazquez2020positive}).
These loci, however, only account for a small part of a pathogen's proteome.
Additionally, these essential parts differ widely between pathogens.
Because of this scarcity and variance in targets, 
one can imagine that the human immune system 
is not tailored to detect these sites, 
as hinted by upon by the aforementioned influenza study.

@@

The effective number of epitopes, however,
is much lower than the actual number of possible
epitopes, due to immunodominance. 
Immunodominance is a feature of the immune system,
due to which some epitopes dominate in causing an immune response,
where other (called subdominant) epitopes end up having no 
effect (\cite{akram2012immunodominance}).
The same study mentions 20 factors that influence the strength
of an immune response.
The most relevant factor for this study is the effect of
a mutation within an epitope, 
such as the in vitro example described in (\cite{berkhoff2004mutation}),
where one single mutation resulted in a 5-20\% weaker
response of virus-specific CD8+ T cells \richel{Fun experiment:
do this in silico}.
Due to immunodominance, we dare assume that there can be selection
for immune system avoidance, as the effective number of loci may be
low and effects of a single mutation noticeable.

@@

When using DNA sequences, one can use a skewed rate 
in non- versus synonymous mutation
to detect the signature of selection (\cite{murrell2015gene}).
Using AA sequences, however, has its advantages,
as its is closer to the actual phenotype
selection acts upon: DNA may never be translated to RNA,
or its RNA may never be transcribed (\cite{diz2012proteomics})
\richel{improve upon this point}.
When using a proteome in phylogenetic research,
we know that the majority of proteins are selected to just 
maintain their function most of the time, where
is the time spans there is selection, only a few AAs
can actually increase the 'fitness' of the 
protein (\cite{anisimova2009investigating}).
There, when generalizing the dynamics of mostly purifying selection (to maintain
a protein's function) and a short duration of positive selection,
those genes that are selected cannot be detected (\cite{yang2000statistical}).

@@

Instead of using the membrane protein topology prediction
tool \verb;TMHMM; (\cite{krogh2001predicting}), PureseqTM (\cite{wang2019efficient})
is used instead.

@@

To detect the evolutionary conservation of TMHs,
we first selected all unique protein sequences per
gene from all 45k SARS-CoV-2 proteomes,
resulting in around 500 unique protein sequences per gene.
From the unique protein sequences, we made an alignment per gene,
using the msa (\cite{bodenhofer2015msa}) R package, 
using default settings, including the use of the default alignment tool
ClustalW (\cite{thompson1994clustal}).
Per alignment, we used msa to determine the consensus sequence
and the conservation scores.
Per consensus sequences, the topology was determined using pureseqtmr.

@@

There are multiple computational tools developed to predict which
parts of membrane proteins are TMH.
In 2001, multiple tools to do so have been compared (\cite{moller2001evaluation}),
of which TMHMM (\cite{krogh2001predicting}) turned out to be the best.
TMHMM is used in the previous study (\cite{bianchi2017}),
yet TMHMM itself has a restrictive software license and is nearly two
decades old. 
Inspired on TMHMM, many other tools followed, such as 
Phobius (\cite{kall2004combined}),
ConPred II (\cite{arai2004conpred}),
MEMSAT3 (\cite{jones2007improving})
and MetaTM (\cite{klammer2009metatm}).
The most recent FOSS tool is called PureseqTM (\cite{wang2019efficient}),
which is used in this study.

@@

For MHC-I, there are multiple computational tools developed 
to predict epitopes. 
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCcons (\cite{karosiene2012netmhccons}) gave the best predictions.
A tool developed later is \verb;epitope-prediction; (\cite{bianchi2017}),
which uses a stabilized matrix method (\cite{kim2009derivation}).
The most recent FOSS tool available is 
MHCnuggets (\cite{shao2020high}) (which can do both MHC-I 
and MHC-II prediction) which is used in this study

@@

Also for MHC-II, there are multiple computational tools developed 
to predict epitopes,
such as using a trained neural network (\cite{nielsen2003reliable})
or a Gibbs sampling approach (\cite{nielsen2004improved}).
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCIIpan (\cite{nielsen2008quantitative,karosiene2013netmhciipan})
gave rise to the most accurate predictions.
Later tools are 
\richel{name of tool here} (\cite{zhang2013predicting}),
\richel{name of tool here} (\cite{trolle2014nettepi}),
\richel{name of tool here} (\cite{zhang2015accurate})
and the very recent MHCnuggets (\cite{shao2020high}),
the latter which is used in this study.

@@

We aimed our evolutionary experiment at TMHs, because these can
be predicted well from a protein structure,
are common structures and are present in all pathogens. 
We could have done the same experiment on beta-turn,
as also these can be predicted well (\cite{petersen2010netturnp}),
are common structures and are present in all pathogens.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster. 
Additionally, we would like to thank Sci-Hub (\cite{himmelstein2018sci})
for allowing us to read paywalled articles while working from home.
